---
title: "Titanic ML"
output: html_document
---

Following along at [this kaggle notebook]("https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic/notebook").


Libraries
```{r}
library(tidyverse)
library(stringr)
library(ggthemes) # visualization
library(scales) # visualization
library(mice) # imputation
library(randomForest) # classification algorithm
library(cowplot) # nice plots
```


Data
```{r}
train <- read.csv("train.csv")
test <- read.csv("test.csv")

full <- bind_rows(train, test) # test data doesn't have susrvived column

str(full)
```


Feature Engineering  

* passenger name has titles and surnames  
    + titles indicate sex position, marriage  
    + surnames indicate family  
    
```{r}
# make colnames lowercase because it's easier to type
colnames(full) <- colnames(full) %>% tolower()

full %>% 
  separate(col = name, into = c('surname', 'fullname'), sep = ", ") %>% 
  separate(col = fullname, into = c('title', 'name'), sep = "[\\.]") -> full

# see table of sex and title counts
table(full$sex, full$title)

# who is the countess and the Jonkheer (Dutch nobility)?
full %>% filter(title == "the Countess") # she survived!
full %>% filter(title == "Jonkheer") # looks like the nobleman died.

```

Titles with low counts combined to "rare"
```{r}
full %>% group_by(title) %>% count() %>% arrange(-n) -> temp
rare_titles <- temp[5:nrow(temp), 1] %>% unlist()

# reassign mlle (unmarried - miss), ms (unmarried - miss), and mme (married - mrs) 
full$title[full$title == "Mlle"] <- "Miss"
full$title[full$title == "Ms"] <- "Miss"
full$title[full$title == "Mme"] <- "Mrs"

# reassign rare titles
full$title[full$title %in% rare_titles] <- "Rare Title"

# show title counts by sex again
table(full$sex, full$title)

```


Do families sink or swim together?
```{r}
unique(full$surname) %>% length() # 875 unique surnames 

# create family size variable including passenger themselves
full$family_size <- full$sibsp + full$parch + 1

# create family variable 
full$family <- paste(full$surname, full$family_size, sep = "_")

unique(full$family) %>% length() # 928 unique families

# visualize
ggplot(full %>% filter(!is.na(survived))) + geom_histogram(aes(family_size, fill = factor(survived)), position = "dodge") + scale_x_continuous(breaks = c(1:max(full$family_size)))

# there's a survival penalty to singles, and families above size 4
```


Discretize family size variable
```{r}
full$family_size_bin[full$family_size == 1] <- "single"
full$family_size_bin[full$family_size < 5 & full$family_size > 1] <- "small"
full$family_size_bin[full$family_size > 4] <- "large"

# show family size by survival using a mosaic plot
mosaicplot(table(full$family_size_bin, full$survived), 
           main='Family Size by Survival',
           ylab = "survived",
           shade=TRUE)
```


Make some more variables
```{r}
full %>% group_by(cabin) %>% count() # lots of missing values

# get deck - first character of cabin variable
full$deck <- substr(full$cabin,1,1)

```


Missingness  

* don't want to delete rows containing missing data  
* so we impute: replace missing values with sensible values given the distirbution of the data:  
    + mean, median, mode
    + prediction  

Find missing values
```{r}
# missing values can be NA, NULL, or "", so write a function that finds missing values
find_missing_values <- function(col_name, df){
  df %>% filter(is.na(get(col_name)) | get(col_name) == "" | is.null(get(col_name)))
}

# apply this function over the column names of our df
col_names <- colnames(full)
missing_vals <- lapply(col_names, find_missing_values, df = full)
names(missing_vals) <- col_names

# explore missing values
str(missing_vals, list.len = 18, max.level = 1)

# survived, age, fare, cabin, embarked, and deck all have missing values. survived is expected, as this reflects our test data
```


Sensible Value Imputation 

We infer where they embarked from by information that we imagine is relevant:  
 
* passenger class  
* fare  
```{r}
# start with embarked because it's small. what passengers don't have info?
missing_vals['embarked'] 

full[c(62, 830), 'embarked'] # passengers 62 and 830 are missing embarked info

# let's look at the rest of the data
full %>% 
  filter(passengerid != 62 & passengerid != 830) %>% 
  ggplot() +
  geom_boxplot(aes(embarked, fare, fill = factor(pclass))) +
  geom_hline(aes(yintercept = 80)) +
  scale_y_continuous(labels = dollar_format())

# the median fare paid by first class passengers departing from Charbourg (C) coincides with the $80 paid by our passengers without embarkment info. Let's replace those "" values with "C"

full$embarked[c(62,830)] <- 'C'

```


Fare missing values-- only 1
```{r}
missing_vals['fare'] 

# this is a 3rd class passenger from Southhampton ('S')

# what did 3rd class passengers from southhampton pay for a fare?

full %>% 
  filter(pclass == 3 & embarked == 'S') %>% 
  ggplot() +
  geom_density(aes(x = fare), fill = "lightblue", alpha = 0.5) + 
  geom_vline(aes(xintercept = median(fare, na.rm = TRUE)), color = "red", linetype = "dashed", lwd = 1) +
  scale_x_continuous(labels = dollar_format())


# we will replace the missing value with the median of this group
fare_med <- full %>% filter(pclass ==3 & embarked == 'S') %>% summarise(median = median(fare, na.rm = TRUE)) # $8.05

full$fare[1044] <- fare_med
```


Predictive Imputation: Lots of missing age data  

create a model that predicts agge from other variables!  
```{r}
sum(is.na(full$age)) # yikes

# this is too much to do by hand.





```






    
    